---
title: "How Does Supermarket Accessibility Vary Across the United States?"
format: 
  html:
    theme: Flatly
    toc: true
    toc-location: right
    toc-depth: 3
    code-fold: true
---

# Executive Summary 

This analysis addresses a primary **research question: How does supermarket accessibility vary across the United States?** This work is one component of a larger collaborative analysis investigating whether "food deserts" are linked to chronic health outcomes, such as diabetes. While my teammates analyzed income levels, fast-food density, and health statistics to build a comprehensive Linear Regression model, my focus is on the spatial distribution of low food access.

By integrating the **USDA Food Access Research Atlas** with the **ACS 2019 Population Dataset**, I map and analyze where the physical distance to quality food is most significant. Finally, I run a logistic regression model to evaluate how effectively socioeconomic factors such as poverty levels, SNAP enrollment, and household vehicle access, can predict whether a census tract is classified as 'Low Access.' This statistical approach allows me to move beyond simple mapping and quantify which factors most significantly influence food insecurity in the U.S.

### Technical Setup
Before importing the datasets, I initiated a standardized workflow by creating a dedicated data directory (if one did not already exist), loading the necessary R libraries, and defining custom functions for title formatting and data cleaning. This ensures the analysis is both reproducible and organized.
```{r}
#| warning: false

library <- function(pkg){
  pkg <- as.character(substitute(pkg))
  options(repos = c(CRAN = "https://cloud.r-project.org"))
  if(!require(pkg, character.only=TRUE, quietly=TRUE)) install.packages(pkg)
  stopifnot(require(pkg, character.only=TRUE, quietly=TRUE))
}

library(tidyverse)
library(stringr)
library(readr) 
library(ggplot2)
library(DT)
library(knitr)
library(kableExtra)
library(scales)
library(broom)

format_titles <- function(df){
  minor_words <- c("and", "of", "the", "in", "by", "for", "to", "or", "a", "an")
  new_names <- str_replace_all(colnames(df), "_", " ") |>
    str_to_title()
  for (word in minor_words) {
    pattern <- paste0("\\b", tools::toTitleCase(word), "\\b")
    replacement <- paste0(word)
    new_names <- str_replace_all(new_names, pattern, replacement)
  }
  colnames(df) <- new_names
  df} 

format_mil <- label_number(accuracy = 0.1, 
                        scale = 1e-6, 
                        suffix = " Million")

# ---- Creating data folder ----
if(!dir.exists(file.path("data", "final"))){
  dir.create(file.path("data", "final"), showWarnings=FALSE, recursive=TRUE)
}
```

# USDA Food Access Research Atlas Dataset 

The USDA [Food Access Research Atlas](https://www.ers.usda.gov/data-products/food-access-research-atlas) (FARA) recognizes that food accessibility is a complex issue influenced by factors like public transportation, household income, and vehicle availability. To account for these variables, the FARA defines **"Low Access" based on the physical distance to the nearest supermarket**. Specifically, a census tract is flagged if at least 500 people, or 33% of the population, live beyond the following thresholds:

- **Half-and-10**: Residents are more than 0.5 miles (Urban) or 10 miles (Rural) from a supermarket.
- **1-and-10**: Residents are more than 1 mile (Urban) or 10 miles (Rural) from a supermarket.
- **1-and-20**: Residents are more than 1 mile (Urban) or 20 miles (Rural) from a supermarket.

The dataset offers two primary lenses: "Low Income and Low Access" (LILA) and general "Low Access." For this analysis, **I focus exclusively on Low Access.** By removing the income constraint from the primary metric, I maintain better control over the variables and isolate the relationship between geography and infrastructure for more accurate findings.

The USDA releases the FARA around every five years, with the most recent being the 2019 FARA. This Atlas aggregated **2019 supermarket data**, **2018 ACS 5-year estimates**, and **2010 U.S. Census population data**. Because of this cycle, my analysis represents a specific "snapshot" in time which is essential to keep in mind as we interpret the results.

### Downloading Dataset
This code downloads and unzips the static file that the USDA offers only if not found locally. 

```{r}
#| warning: false

FOOD_FILENAME <- file.path("data", "final", "2019 Food Access Research Atlas Data.zip")

if(!file.exists(FOOD_FILENAME)){
  download.file("https://ers.usda.gov/sites/default/files/_laserfiche/DataFiles/80591/2019%20Food%20Access%20Research%20Atlas%20Data.zip", 
                destfile=FOOD_FILENAME)
}

if(!file.exists(FOOD_FILENAME)){
  unzip(FOOD_FILENAME, exdir = file.path("data", "final"))
}

FOOD_ATLAS_FILENAME <- file.path("data", "final", "Food Access Research Atlas.csv")
```

### Data Cleaning Notes

```{r}
#| warning: false

FOOD_ATLAS <- read_csv(FOOD_ATLAS_FILENAME) |>
  select(
    "CensusTract", "State", "County", "Urban",
    "LowIncomeTracts", "LA1and10", "LAhalfand10", "LA1and20",
    "PovertyRate",  "OHU2010", "TractHUNV", "TractSNAP", "Pop2010",
    "LAPOP1_10","LAPOP05_10","LAPOP1_20"
  )

missing_zero_tracts <- FOOD_ATLAS |>
  mutate(geoid_length = str_length(CensusTract))|>
  count(geoid_length) |>
  filter(geoid_length == 10) |>
  pull(n)

FOOD_ATLAS <- FOOD_ATLAS |>
  mutate(
    CensusTract = str_pad(
      string = as.character(CensusTract),
      width = 11,
      pad = "0",
      side = "left"
    )
  )

# Checking the that all tracts are 11 digits
#FOOD_ATLAS |>
  #mutate(geoid_length = str_length(CensusTract)) |>
  #count(geoid_length, sort = TRUE)

FOOD_ATLAS <- FOOD_ATLAS |>
  mutate(LAPOP1_10 = as.numeric(LAPOP1_10),
         LAPOP05_10 = as.numeric(LAPOP05_10),
         LAPOP1_20 = as.numeric(LAPOP1_20),
         LAPOP1_10 = replace_na(LAPOP1_10, 0),
         LAPOP05_10 = replace_na(LAPOP05_10, 0),
         LAPOP1_20 = replace_na(LAPOP1_20, 0))
```
To ensure the FARA dataset is properly structured for efficient analysis and accurate merging with American Community Survey population data, the following steps were taken:

- **Column Selection**: only selected the necessary 16 columns from the original 47 to create a more focused workspace.

- **GEOID Standardization**: The `CensusTract` column is standardized to the required 11-digit (FIPS) code format. A total of **`r comma(missing_zero_tracts)`** tracts are padded with a leading zero to ensure accurate spatial joining with the ACS population data. 

- **Data Type and Missing Value**: The Low Access Population count variables (`LAPOP1_10`, `LAPOP05_10`, `LAPOP1_20`) are converted to numeric data types. Additionally, all NA were replaced with 0 values. 

### FARA Dataset Preview
::: {.callout-note}
## USDA FARA Variable Dictionary
`Urban`, `Low Income`, `1 and 10`, `Half and 10`, and `1 and 20`  are dummy variables, where a value of 1 signifies that the census tract meets the stated criteria (Yes), and a value of 0 signifies that it does not (No).
```{r dictionary_table}
# 1. Create the data frame from your dictionary
dictionary_df <- tribble(
  ~`Column Name`, ~Description,
  "Urban", "Flag for urban census tract.",
  "Low Income", "Flag for low income census tract.",
  "1 and 10", "Flag for low access tract (1 mile for urban, 10 miles for rural).",
  "Half and 10", "Flag for low access tract (1/2 mile for urban, 10 miles for rural).",
  "1 and 20", "Flag for low access tract (1 mile for urban, 20 miles for rural).",
  "Poverty %", "Share of the tract population living with income at or below the Federal poverty thresholds.",
  "OHU2010", "Occupied housing unit count from 2010 census.", 
  "TractHUNV", "Total count of housing units without a vehicle in the tract.",
  "TractSNAP", "Total count of housing units receiving SNAP benefits in the tract.",
  "Pop2010", "Population count from the 2010 census.",
  "Pop 1 and 10", "Population count beyond 1 mile (urban) or 10 miles (rural) from a supermarket.",
  "Pop Half and 10", "Population count beyond 1/2 mile (urban) or 10 miles (rural) from a supermarket.",
  "Pop 1 and 20", "Population count beyond 1 mile (urban) or 20 miles (rural) from a supermarket."
)

# 2. Use kable to format the table with specified column widths
dictionary_df |> 
  mutate(`Column Name` = cell_spec(`Column Name`, bold = TRUE)) |> # Bolds the first column
  kable(
    format = "html", 
    escape = FALSE, # Allows HTML for bolding
    col.names = c("Column Name", "Description"),
    # Set the width of the columns! 20% for the first, 80% for the second.
    col.widths = c("20%", "80%") 
  )
```
:::

```{r}
FOOD_ATLAS |>
  rename("Census Tract" = "CensusTract",
         "Low Income" = "LowIncomeTracts",
         "1 and 10" = "LA1and10",
         "Half and 10" = "LAhalfand10", 
         "1 and 20" = "LA1and20",
         "Poverty %" = "PovertyRate",
         "Pop 1 and 10" = "LAPOP1_10",
         "Pop Half and 10" = "LAPOP05_10",
         "Pop 1 and 20" = "LAPOP1_20") |>
  head(n =10) |> 
  datatable(options=list(searching=FALSE, info=FALSE, dom = 'tip', pageLength = 5)) |>
  formatRound(c("Pop2010", "Pop 1 and 10","Pop Half and 10",  "Pop 1 and 20"), digits = 0)
```

# U.S Census ACS Population Dataset

The U.S. Census Bureau’s [American Community Survey](https://www.census.gov/programs-surveys/acs/) (ACS) provides granular demographic, economic, and housing data by surveying a random sample of 3.5 million households annually. For this analysis, I imported the ACS [2019 Total Population](https://data.census.gov/table/ACSDT5Y2019.B01003?q=Population+Total&g=010XX00US$1400000) (B01003) 5-year estimates to align with the 2019 supermarket data.
While the 2019 USDA FARA includes population figures, those metrics are based on the 2010 Decennial Census. I chose to integrate the 2019 ACS data to explore how more contemporary population counts compare to supermarket distances.

::: {.callout-important}
To maintain the integrity of the original FARA "Low Access" ratios, all subsequent analysis and statistics will utilize the 2010 Census Population unless explicitly stated otherwise. This ensures my findings remain technically consistent with the USDA’s official flagging criteria.
:::

### Downloading Dataset

To acquire the population data, I utilized the `tidycensus` package to interface with the U.S. Census Bureau API through a conditional pipeline. To optimize performance, the script is designed to only download the 2019 ACS 5-year estimates if the local RDS file is missing. Recognizing that this is a large dataset, I also included a pre-flight check: if a user lacks a Census API key, a conditional statement prints clear instructions on how to retrieve and install one, ensuring the code is accessible to other users.

To import census tracts for the entire country, I implemented a loop that iterates through all state FIPS codes, specifically excluding U.S. territories to keep the analysis focused on the 50 states and D.C.. Once the loop completes, I perform a relational join to merge the demographic population counts with the spatial geometry, creating a unified dataset ready for geographic analysis.

```{r}
#| message: false
#| warning: false


library(tidycensus)
library(tigris)
library(sf)
library(tidyr)

# Census API Key (only needed once)
#census_api_key("1885fa1a37974bd72e037147e7e0b715bbcbfd37", install = TRUE)

if (Sys.getenv("CENSUS_API_KEY") == "") {
  stop("CENSUS_API_KEY environment variable not found.
        Please obtain a Census API key from 
        https://api.census.gov/data/key_signup.html and then run:
        tidycensus::census_api_key('YOUR_KEY_HERE', install = TRUE)
        ", call. = FALSE)
}

POPULATION_FILENAME <- file.path("data", "final", "us_census_population_geo.rds")

if (!file.exists(POPULATION_FILENAME)) {
  
  message("File not found locally. Starting download and processing...")
  
  options(tigris_use_cache = TRUE)
  
  fips_codes <- unique(fips_codes$state)[1:51]
  fips_codes <- fips_codes[!(fips_codes %in% c("72", "78", "60", "66", "69"))]
  
  all_tract_data_list <- list()
  all_tract_geometry_list <- list()
  
  # Loop to download both ACS data and Geometry
  for (fips in fips_codes) {
    all_tract_data_list[[fips]] <- get_acs(
      geography = "tract",
      table = "B01003",
      state = fips,
      year = 2019,
      output = "wide"
    )
    
    all_tract_geometry_list[[fips]] <- tracts(
      state = fips,
      year = 2019,
      class = "sf"
    )
  }
  
  # Combine Data and Geometry
  POPULATION <- do.call(rbind, all_tract_data_list)
  TRACT_GEOMETRY <- do.call(rbind, all_tract_geometry_list)
  
  # Cleaning
  POPULATION <- POPULATION |>
    separate(
      col = NAME,
      into = c("CensusTract", "County", "State"),
      sep = ", "
    ) |>
    rename("Population" = "B01003_001E") |>
    select(-"B01003_001M") |>
    filter(Population != "0") |>
    mutate(CensusTract = str_replace(CensusTract, "Census Tract ", ""))

  TRACT_GEOMETRY <- TRACT_GEOMETRY |>
    select(GEOID, geometry)

  POPULATION_GEO <- POPULATION |>
    left_join(TRACT_GEOMETRY, by = "GEOID") |>
    st_as_sf()
  
  # --- save the file ---
  saveRDS(POPULATION_GEO, file = POPULATION_FILENAME)
  message("Download, processing, and saving complete.")
  
} else {
  
  # --- Load the saved dataset if the file EXISTS ---
  POPULATION_GEO <- readRDS(POPULATION_FILENAME)
  message("Dataset loaded from local file (", basename(POPULATION_FILENAME), "). Skipping download.")
}
```

### Data Cleaning Notes
The following steps were taken to process the ACS population data obtained via `tidycensus` :

- **Column Renaming and Selection**: The variable `B01003_001E` (estimate total population) is renamed to Population and `B01003_001M`(margin of error)is removed since it is not required for this project.
- **Column Separation**: The initial `NAME` column, which combines the census tract, county, and state, is separated using the comma delimiter (, ) to isolate and correctly structure the `CensusTract`, `County`, and `State`.
- **Tract Filtering**: To ensure a relevant population base for per-capita analysis, any census tract with a reported Population count of 0 are filtered out of the data set.
- **Territory Filtering**: The analysis focuses on the U.S. states and the District of Columbia. The following American territories, Puerto Rico, U.S. Virgin Islands, American Samoa, Guam, and Northern Mariana Islands, are filtered out. Further analysis on these territories can be interesting future work!

### Population Dataset Preview

```{r}
POPULATION_GEO |>
  st_drop_geometry() |>
  rename("Census Tract" = "CensusTract") |>
  head(n =10) |> 
  datatable(options=list(searching=FALSE, info=FALSE, dom = 'tip', 
                         paging = FALSE),
            rownames = FALSE) |>
  formatRound("Population", digits = 0)
```

# Merge Data 

To ensure that Alaska and Hawaii are clearly visible without distorting the scale of the continental choropleth maps, I utilized the `shift_geometry()` function. Additionally, I transformed the coordinates to the **WGS84** (World Geodetic System 1984) standard. This step ensures spatial consistency across both datasets and allows for a more cohesive visual representation of food access across all fifty states.
```{r}
FOOD_POP <- POPULATION_GEO |>
  inner_join(
    FOOD_ATLAS, by = c("GEOID" = "CensusTract",
                       "State" = "State",
                       "County" = "County"))

FOOD_POP <- FOOD_POP |>
  shift_geometry() |> 
  st_transform(crs="WGS84") 

FP_NOGEO <- FOOD_POP |>
  st_drop_geometry() 
```

### Final Data Cleaning 
```{r}
#total number of tracts 
total_tracts <- FP_NOGEO |>
  count()|>
  pull(n)

#checking for duplicates
duplicates <- FP_NOGEO |> 
  duplicated() |>
  sum()

low_pop <- FP_NOGEO |>
  filter(Pop2010 <= 1200) |>
  arrange(Pop2010)

special_tracts_count  <- FP_NOGEO |>
  filter(str_starts(CensusTract, "98") | str_starts(CensusTract, "99")) |>
  count()|>
  pull(n)

FOOD_POP <- FOOD_POP |>
  filter(!( (Urban == 1 & Population < 1200) | (Population < 20)))

FP_NOGEO <- FOOD_POP |>
  st_drop_geometry() 

#final number of tracts
final_tract_count <- FP_NOGEO |>
  count()|>
  pull(n)

```

The following steps finalize the preparation of the final merged dataset:

- Total tracts before final edits: **`r comma(total_tracts)`** 
- **Filtering Low Population**: While tracts with zero population are filtered out during the import phase, further filtering is required to remove low-population outliers. Based on U.S. Census Bureau, the population for standard census tracts ideally ranges from 1,200 to 8,000, with an optimal target of 4,000. Therefore, the following tracts were filtered out:
    - Tracts flagged as **urban** and having a population **less than 1,200** people.
    - Tracts with a **less than 20 people** (regardless of urban status).
- Total tracts after edits: **`r comma(final_tract_count)`** 

### Final Merged Dataset Preview 
```{r}
FP_NOGEO |>
  rename("Census Tract" = "CensusTract",
         "Low Income" = "LowIncomeTracts",
         "1 and 10" = "LA1and10",
         "Half and 10" = "LAhalfand10", 
         "1 and 20" = "LA1and20",
         "Poverty %" = "PovertyRate",
         "Pop 1 and 10" = "LAPOP1_10",
         "Pop Half and 10" = "LAPOP05_10",
         "Pop 1 and 20" = "LAPOP1_20") |>
  head(n =10) |> 
  datatable(options=list(searching=FALSE, info=FALSE, dom = 'tip', pageLength = 5), 
            rownames = FALSE) |>
  formatRound(c("Pop2010", "Pop 1 and 10","Pop Half and 10", "Pop 1 and 20","Population"), digits = 0)

```

# Where are the low access tracts in the US?
The data is now ready for analysis! To kick off the EDA, I will first visualize how low food access varies across the United States. 

### Interactive Low Access Map
To provide a more dynamic view of the data, I developed an interactive Leaflet map that allows users to toggle between the three primary distance metrics: **LA1and10, LAHalfand10, and LA1and20**. The goal of this visualization is to allow for rapid comparison of how each definition changes the geographic landscape of low access in real-time. This interactive feature is especially valuable when comparing the 1-and-10 and Half-and-10 metrics, where the subtle differences in tract flagging are most apparent when zooming in at the state or local level.

```{r}
#| echo: true
#| message: false
#| warning: false


library(leaflet)
library(sf)
library(rmapshaper) 

#---- Preparing Census Tract borders ----
options(tigris_use_cache = TRUE) 

# Download Census Tracts (2019)
us_tracts <- tracts(year = 2019,class = "sf", cb = TRUE) |>
  filter(!STATEFP %in% c("02", "15")) |>
  mutate(GEOID = as.character(GEOID)) |>
  st_transform(crs="WGS84") |>
  select(GEOID)

#letting me run without my laptop crashing
us_tracts_simple <- ms_simplify(us_tracts , keep = 0.1, keep_shapes = TRUE)

# General check 1: making sure geoid are formatted properly
us_tracts_digits <- us_tracts |>
  st_drop_geometry()|>
  mutate(geoid_length = str_length(GEOID)) |>
  count(geoid_length, sort = TRUE)

#joining geo with food_pop
leaflet_data <- us_tracts_simple |>
  inner_join(FP_NOGEO, by = "GEOID")

#checking for duplicates
#leaflet_data |> st_drop_geometry()|> duplicated() |>sum()

interactive_data <- leaflet_data |>
  mutate(
    label_1and10 = case_when(
      LA1and10 == 1 & Urban == 1 ~ "Urban Low Access",
      LA1and10 == 1 & Urban == 0 ~ "Rural Low Access",
      TRUE ~ "No"
    ),
    label_half = case_when(
      LAhalfand10 == 1 & Urban == 1 ~ "Urban Low Access",
      LAhalfand10 == 1 & Urban == 0 ~ "Rural Low Access",
      TRUE ~ "No"
    ),
    label_1and20 = case_when(
      LA1and20 == 1 & Urban == 1 ~ "Urban Low Access",
      LA1and20 == 1 & Urban == 0 ~ "Rural Low Access",
      TRUE ~ "No"
    )
  )

factpal <- colorFactor(
  palette = c("Urban Low Access" = "#932092", "Rural Low Access" = "#0432ff"),
  domain = c("Urban Low Access", "Rural Low Access")
)

# Map with Subsets
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  
  # Layer 1: 1 and 10 (Filtered)
  addPolygons(
    data = subset(interactive_data, label_1and10 != "No"), 
    group = "LA 1 and 10",
    fillColor = ~factpal(label_1and10),
    fillOpacity = 0.7,
    color = NA,      # remove borders for speed
    popup = ~paste0("State: ", State, 
                    "<br>County: ", County, 
                    "<br>Tract: ", CensusTract, 
                    "<br>Status: ", label_1and10)
  ) |>
  
  # Layer 2: Half and 10 (Filtered)
  addPolygons(
    data = subset(interactive_data, label_half != "No"),
    group = "LA Half and 10",
    fillColor = ~factpal(label_half),
    fillOpacity = 0.8,
    color = NA,
    popup = ~paste0("State: ", State, 
                    "<br>County: ", County, 
                    "<br>Tract: ", CensusTract, 
                    "<br>Status: ", label_1and10)
  ) |>
  
  # Layer 3: 1 and 20 (Filtered)
  addPolygons(
    data = subset(interactive_data, label_1and20 != "No"),
    group = "LA 1 and 20",
    fillColor = ~factpal(label_1and20),
    fillOpacity = 0.7,
    color = NA,
    popup = ~paste0("State: ", State, 
                    "<br>County: ", County, 
                    "<br>Tract: ", CensusTract, 
                    "<br>Status: ", label_1and10)
  ) |>
  
  addLayersControl(
    baseGroups = c("LA 1 and 10", "LA Half and 10", "LA 1 and 20"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addLegend(
    pal = factpal,
    values = c("Urban Low Access", "Rural Low Access"),
    title = "Access Status",
    position = "bottomright"
  )
```
**Reflections**: During the development phase, I encountered significant performance issues when attempting to render the full-resolution geometry from the ACS population import. To prevent system crashes and ensure a smooth user experience, I opted to redownload the tract boundaries and simplify the geometries to reduce the computational load. For future projects, a more efficient approach would be to import and manage population data and spatial geometries as separate components from the start.

### Number of Low Access Tracts by State
*Note: `Total Tracts` represent the total number of tracts in the state.*
```{r}
low_access_tracts_by_state <- FP_NOGEO |>
  group_by(State) |>
  summarise(
    total_half_and_10 = sum(LAhalfand10),
    total_1_and_10 = sum(LA1and10),
    total_1_and_20 = sum(LA1and20),
    total_tracts = n()
  ) |>
  ungroup() |>
  arrange(desc(total_1_and_10))

low_access_tracts_by_state |>
  rename("0.5 / 10 mi" = "total_half_and_10",
         "1 / 10 mi" = "total_1_and_10",
         "1 / 20 mi" = "total_1_and_20", 
         "Total Tracts" = "total_tracts") |>
  datatable(options=list(searching=FALSE, info=FALSE, dom = 'tip'), 
            rownames = FALSE) |>
  formatRound(c("0.5 / 10 mi", "1 / 10 mi","1 / 20 mi", "Total Tracts"), digits = 0)

```
**Findings**: Based on the 1 and 10 metric, the total count of low access census tracts is **highest in Texas, California, and Florida**, and **lowest in Vermont, Wyoming, and Alaska**.

### Percent of Low Access Tracts in each State
```{r}
low_access_tracts_by_state_percent <- low_access_tracts_by_state |>
  group_by(State) |>
  summarise(
    percent_LAhalfand10 = (total_half_and_10 / total_tracts),
    percent_LA1and10 = (total_1_and_10 / total_tracts),
    percent_LA1and20 = (total_1_and_20 / total_tracts)
    ) |>
  ungroup() |>
  arrange(desc(percent_LA1and10)) 
  
low_access_tracts_by_state_percent |>
  rename("0.5 / 10 mi" = "percent_LAhalfand10",
         "1 / 10 mi" = "percent_LA1and10",
         "1 / 20 mi" = "percent_LA1and20") |>
  datatable(options=list(searching=FALSE, info=FALSE, dom = 'tip'), 
            rownames = FALSE)|>
  formatPercentage(c("0.5 / 10 mi", "1 / 10 mi","1 / 20 mi"), digits = 2)
  
```
**Findings**: Based on the 1 and 10 metric, the percent of low access census tracts is **highest in Wyoming, South Dakota, and Mississippi**, and **lowest in New York, Vermont, and Maine**. 

# How many people live in low access areas?

With a broad understanding of the location and concentration of low-access tracts at the state level, we can now examine the populations residing within these areas.

```{r}
total_pop110 <- FP_NOGEO |>
  summarise(total = sum(LAPOP1_10)) |>
  pull(total)

total_pop0510 <- FP_NOGEO |>
  summarise(total = sum(LAPOP05_10))|>
  pull(total)

total_pop120 <- FP_NOGEO |>
  summarise(total = sum(LAPOP1_20))|>
  pull(total)
```
Even in tracts that do not meet the low access metric threshold to be labeled a "food desert," millions of individuals still live beyond reasonable distances from a supermarket. The following are the total approximate US population residing in low access areas: 

- Half-and-10: **`r format_mil(total_pop0510)`**
- 1-and-10: **`r format_mil(total_pop110)`**
- 1-and-20: **`r format_mil(total_pop120)`**


### State Level Low Access: Population
*Note: `Total Population` represent the total population of the state, not an aggreated population total of low access.*
```{r}
LA_POP <- FP_NOGEO |>
  group_by(State) |>
  summarise(
    total_half_and_10 = sum(LAPOP05_10),
    total_1_and_10 = sum(LAPOP1_10),
    total_1_and_20 = sum(LAPOP1_20),
    total_population = sum(Pop2010)
  ) |>
  ungroup() |>
  arrange(desc(total_1_and_10))

LA_POP |>
  rename("0.5 / 10 mi" = "total_half_and_10",
         "1 / 10 mi" = "total_1_and_10",
         "1 / 20 mi" = "total_1_and_20",
         "Total Population" = "total_population") |>
  datatable(options=list(searching=FALSE, info=FALSE, dom = 'tip'), 
            rownames = FALSE)|>
  formatRound(c("0.5 / 10 mi", "1 / 10 mi","1 / 20 mi", "Total Population"), digits = 0)

```
**Findings**: The population living in low-access areas (using the 1 and 10 metric) is **highest in Texas, California, and Florida**, and **lowest in Vermont, Wyoming, and Maine**. 

### Visual: State Level Low Access: Population
Visualizing population totals by state highlights **Texas** as the most heavily impacted region, with a low-access population nearly triple that of most other states.
```{r}
#| message: false
#| results: hide

us_states <- states(cb = TRUE, resolution = "20m") |>
  shift_geometry() |> # Automatically moves AK/HI below the mainland
  filter(NAME != "Puerto Rico") |>
  select(NAME, geometry)

state_map_pop <- us_states |>
  left_join(LA_POP, by = c("NAME" = "State"))

ggplot(state_map_pop) +
  geom_sf(aes(fill = total_1_and_10), color = "white", size = 0.2) +
  scale_fill_gradient(
    low = "#f1f1f1",        
    high = "#932092",  
    trans = "sqrt",
    labels = scales::label_number(suffix = "M", scale = 1e-6),
    guide = guide_colorbar(
      title = "Total Residents (Square Root Transformed)", 
      title.position = "top", 
      title.hjust = 0.5,
      barwidth = 9,
      barheight = 0.5
    ) 
  ) +
  theme_void() +
  labs(
    title = "Total Population Residing in Low Food\nAccess Areas by State",
    subtitle = "Metric: 1 mile for Urban / 10 miles for Rural",
    caption = "Source: USDA Food Access Research Atlas"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic")
  )
```


### State Level Low Access: Percentage of Population
```{r}
#| message: false
LA_POP_PERCENT <- LA_POP |>
  group_by(State) |>
  summarise(
    percent_halfand10 = (total_half_and_10 / total_population),
    percent_1and10 = (total_1_and_10 / total_population),
    percent_1and20 = (total_1_and_20 / total_population)
  ) |>
  ungroup() |>
  arrange(desc(percent_1and10)) 

LA_POP_PERCENT |>
  rename("0.5 / 10 mi" = "percent_halfand10",
         "1 / 10 mi" = "percent_1and10",
         "1 / 20 mi" = "percent_1and20") |>
  datatable(options=list(searching=FALSE, info=FALSE, dom = 'tip'), 
            rownames = FALSE)|>
  formatPercentage(c("0.5 / 10 mi", "1 / 10 mi","1 / 20 mi"), digits = 2)
```
**Findings**: The percent of population living in low-access areas (using the 1 and 10 metric) is **highest in New Mexico, Alaska, and Georgia**, and **lowest in Vermont, New York, and California**. 


### Visual: State Level Low Access: Percentage of Population
Unlike the population choropleth, this visualization give us a standardized view of how percent of population compares among the 50 states. Notably, while **California** had the second highest population living in low access, it has the third lowest percent of population. 
```{r}
#| message: false
#| results: hide

state_map_data <- us_states |>
  left_join(LA_POP_PERCENT, by = c("NAME" = "State"))

ggplot(state_map_data) +
  geom_sf(aes(fill = percent_1and10), color = "white", size = 0.2) +
  scale_fill_gradient(
    low = "#f1f1f1",        
    high = "#0432ff",          
    labels = scales::label_percent(accuracy = 1), 
    name = "% Low Access"
  ) +
  theme_void() +
  labs(
    title = "Percentage of Population with Low Food\nAccess by State",
    subtitle = "Metric: 1 mile for Urban / 10 miles for Rural",
    caption = "Source: USDA Food Access Research Atlas"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic")
  )
```

# How does population change over time in low access tracts?

To capture a more contemporary snapshot of the residents living within low-access areas, this choropleth utilizes **ACS 2019 5-Year Estimate population** figures. It is important to reiterate that being located within a "low-access tract" does not imply the entire population lives with low food access; rather, this visual serves to highlight the scale of the potential challenge.

```{r}
us_low_access_rural <- FOOD_POP |>
  filter(Urban == 0) |>
  mutate(
    LowAccessPop2010 = case_when(
      LA1and20 == 1 | LA1and10 == 1 ~ Pop2010, 
      TRUE ~ NA_real_ 
    ),
    LowAccessPop2019 = case_when(
      LA1and20 == 1 | LA1and10 == 1 ~ Population, 
      TRUE ~ NA_real_ 
    ),
    netchange = LowAccessPop2019 - LowAccessPop2010
  )

ggplot(us_low_access_rural) +
  geom_sf(aes(fill = LowAccessPop2019), color = NA) +
  scale_fill_gradient(
    low = "white",
    high = "#0432ff",
    trans = "sqrt", 
    na.value = "grey90",
    guide = guide_colorbar(
      title = "Population per Tract (Square Root Transformed)", 
      title.position = "top", 
      title.hjust = 0.5,
      title.size = 10,
      barwidth = 9,
      barheight = 0.5
    ) 
  ) +
  theme_void() +
  labs(
    title = "Total Population within Rural Low Food Access Tracts",
    subtitle = "2019 ACS census tract population for areas defined by the 1-10 or 1-20 mile low access metric",
    caption = "Source: USDA Food Access Research Atlas (2019),\nU.S. Census Bureau ACS B01003 for 2019"
  ) +
  theme(legend.position = 'bottom',
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0, size = 8, face = "italic"))
```
**Findings**: The map reveals a high concentration of low access rural tracts across the **Western United States** and the **Great Plains**, where vast land areas naturally create significant distance barriers for residents. Additionally not only are these tracts extremely large in geographic size but also contain lower than average population densities. These factors combine to create a unique regional landscape where the number of supermarkets are not as near compared to urban areas. 


### Net Change in Population in Rural Low Access Tracts

By comparing the 2010 Census to the 2019 ACS Estimates, we can identify regions where the population has either intensified or receded.
```{r}

ggplot(us_low_access_rural) +
  geom_sf(aes(fill = netchange), color = NA) +
  scale_fill_gradient2(
    low = "#fd9c58",        # Orange for population loss
    mid = "white",          # White for little to no change
    high = "#932092",       # Purple for population gain
    midpoint = 0,           # Critical: ensures colors split at zero
    na.value = "grey85",
    limits = c(-1500, 1500),
    oob = scales::squish,
    labels = scales::label_comma(),
    guide = guide_colorbar(
      title = "Net Population Change", 
      title.position = "left", 
      title.hjust = 0.5,
      barwidth = 11,
      barheight = 0.5
    ) 
  ) +
  theme_void() +
  labs(
    title = "Population Shifts in Rural Low-Access Tracts",
    subtitle = "Net change in residents (2010 Census to 2019 ACS Estimates)",
    caption = "Data Source: USDA Food Access Research Atlas (2019)"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0, size = 8, face = "italic")
  )

```
Because the majority of net changes fell within a range of **+/- 3,000 residents**, with the exception of one outlier of 4,000, I utilized the `scales::squish` function to manually adjust the legend’s range. This increased the visual contrast, allowing for a clearer identification of demographic trends.

The resulting visualization reveals that many of the deep purple tracts, indicating growth, are smaller in geographic size, while the orange-shaded tracts, indicating decline, are significantly larger. This spatial pattern potentially indicates a trend where residents are moving toward more densely populated areas over time, even within areas characterized by low food access.


### Change of Total Population in Rural Low Access Tracts
```{r}
rural_trend_summary <- us_low_access_rural |>
  st_drop_geometry() |>
  summarise(
    `Total Population (2010)` = sum(as.numeric(LowAccessPop2010), na.rm = TRUE),
    `Total Population (2019)` = sum(as.numeric(LowAccessPop2019), na.rm = TRUE),
    `Net Change` = `Total Population (2019)` - `Total Population (2010)`,
    `Percent Growth` = (`Net Change` / `Total Population (2010)`) * 100
  )

kable(rural_trend_summary, digits = 1, format.args = list(big.mark = ","), caption = "Population Growth within Rural Low-Access Tracts (2010-2019)")
```
**Findings**: The overall growth rate of -0.5% confirms the minimal population shift within rural low access tracts, as shown in the Population Shift visual.   

# How does SNAP and vehicle access vary across the US?

### Households with No Vehicle Distribution
```{r}
#| warning: false
hunv_dist <- FP_NOGEO |>
  filter(OHU2010 >= TractHUNV, OHU2010 > 0) |>
  mutate(hunv_percent = (TractHUNV / OHU2010) * 100) |>
  select(GEOID, State, County, OHU2010, TractHUNV, hunv_percent) |>
  filter(!is.na(hunv_percent))

hunv_mean <- hunv_dist |>
  summarise(mean = mean(hunv_percent, na.rm = TRUE)) |>
  pull(mean)

ggplot(hunv_dist, aes(x = hunv_percent)) +
  geom_histogram(
    binwidth = 2,           
    fill = "#b4e6a2",       
    color = "white",       
    alpha = 1
  ) + 
  coord_cartesian(xlim = c(0, 50)) +
  theme_minimal() +
  labs(
    title = "Distribution of Households with No Vehicle by Census Tract",
    subtitle = "Percentage of occupied housing units with no vehicle available per census tract",
    x = "Percent of HUNV",
    y = "Number of Census Tracts",
    caption = "Source: USDA Food Access Research Atlas"
  ) +
  geom_vline(
    aes(xintercept = hunv_mean), 
    color = "#0432ff", linetype = "dashed", size = 1, alpha = 0.6
  ) +
  annotate(
    "text", 
    x = hunv_mean + 9, 
    y = 12000, 
    label = paste0("National Average: ", round(hunv_mean, 2), "%"),
    color = "#0432ff"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    plot.subtitle = element_text(hjust = 0.5, size = 7),
    plot.caption = element_text(hjust = 0, size = 12, face = "italic")
  )
```

```{r}
hunv_summary <- summary(hunv_dist$hunv_percent) |> as.array() |> t()

kable(hunv_summary, caption = "Summary Statistics for No Vehicle Access(%)", digits = 2)
```
**Findings**: The distribution of households without vehicle access is heavily right-skewed, revealing that while most households own a vehicle, a small number of communities are almost entirely dependent on alternative transportation. 

### SNAP Participation Distribution
```{r}
#| warning: false

# Normalize SNAP numbers 
snap_dist <- FP_NOGEO |>
  filter(OHU2010 >= TractSNAP, OHU2010 > 0) |>
  mutate(snap_percent = (TractSNAP / OHU2010) * 100) |>
  select(GEOID, State, County, OHU2010, TractSNAP, snap_percent) |>
  filter(!is.na(snap_percent))

snap_mean <- snap_dist |>
  summarise(mean = mean(snap_percent, na.rm = TRUE)) |>
  pull(mean)

ggplot(snap_dist, aes(x = snap_percent)) +
  geom_histogram(
    binwidth = 2,          
    fill = "#e792e7",      
    color = "white",        
    alpha = 0.8
  ) + 
  coord_cartesian(xlim = c(0, 65)) +
  theme_minimal() +
  labs(
    title = "Distribution of SNAP Participation by Census Tract",
    subtitle = "Percentage of households receiving SNAP benefits",
    x = "Percent of Households on SNAP",
    y = "Number of Census Tracts",
    caption = "Source: USDA Food Access Research Atlas"
  ) +
  geom_vline(
    aes(xintercept = snap_mean), 
    color = "#0432ff", linetype = "dashed", size = 1, alpha = 0.6
  ) +
  annotate(
    "text", 
    x = snap_mean + 12, 
    y = 6000, 
    label = paste0("National Average: ", round(snap_mean, 2), "%"),
    color = "#0432ff"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0, size = 9, face = "italic")
  )
```

```{r}
snap_summary <- summary(snap_dist$snap_percent) |> as.array() |> t()

kable(snap_summary, caption = "Summary Statistics for SNAP Participation (%)", digits = 2)
```
**Findings**: Similar to the vehicle access data, the distribution of SNAP participation is notably right-skewed, indicating that while many census tracts have low levels of enrollment, there are significant outliers with very high participation rates.

# Logistic Regression

To bring the entire analysis together, I ran a logistic regression model to investigate how poverty rates, vehicle access, and SNAP participation influence the likelihood of a tract being classified as low access. To ensure a more accurate comparison across different community sizes, I converted the raw counts of no-vehicle households and SNAP recipients into percentages. 

```{r}
#| warning: false
model_data <- FP_NOGEO |>
  select(LA1and10, PovertyRate, TractSNAP, TractHUNV, OHU2010) |>
  mutate(snap_percent = (TractSNAP / OHU2010) * 100) |>
  mutate(hunv_percent = (TractHUNV / OHU2010) * 100) |>
  drop_na()

log_model <- glm(
  LA1and10 ~ PovertyRate + snap_percent + hunv_percent, 
  data = model_data, 
  family = "binomial"
)

log_results <- tidy(log_model, exponentiate = TRUE, conf.int = TRUE)

kable(
  log_results, 
  digits = 3, 
  caption = "Logistic Regression Results: Odds Ratios for Low Food Access (1 & 10 Miles)"
)
```

### Overall Findings

The logistic regression model confirms that socioeconomic factors are significant predictors of food access, though not always in the expected direction. **SNAP participation has a positive relationship with low access** (OR = 1.006, p < 0.001), meaning a 1% increase in enrollment slightly raises the likelihood of a tract being classified as low access. Conversely, **vehicle access shows a significant negative relationship** (OR = 0.938, p < 0.001). This is likely due to the high proximity of stores in dense, transit-heavy urban areas where residents may not own cars but are still physically close to supermarkets.

Surprisingly, **Poverty Rate was not a significant predictor in this model** (p = 0.099). This suggests that there is more to analyze to better understand this unexpected result. For instance, a future study separating urban and rural tracts might reveal clearer, more localized findings. Additionally, further research into the specific distance metrics used to define food access could provide a more nuanced and accurate picture of how different populations experience food insecurity.

# Final Summary Report: Linear Regression 

### Downloading and Merging Team's Data
```{r}
#| message: false
#| warning: false
merged <- FP_NOGEO |>
  select(GEOID, LA1and10, Population)

#---- NANDA ----

FAST_FILENAME <- file.path("data", "final", "nanda_eatdrink_Tract20_1990-2021_01P.csv")

FAST <- read_csv(FAST_FILENAME) |>
  filter(year == 2019) |> 
  arrange(desc(count_fastfood)) |> 
  rename('GEOID' = 'tract_fips20') |>
  select(GEOID, count_fastfood)

merged <- merged |>
  left_join(FAST, by = "GEOID")

merged <- merged |>
  mutate(count_fastfood = replace_na(count_fastfood, 0)) |>
  mutate(den_fastfood = (count_fastfood / Population) * 1000) |>
  select(-count_fastfood, -Population)

#----INCOME----  

library(readxl)
#for local
INCOME_FILENAME <- file.path("data", "final","ACSDT5Y2018.B19013-Data.xlsm")

INCOME <- read_excel(INCOME_FILENAME)

INCOME <- INCOME |>
  rename('GEOID' = 'tract')|>
  mutate(med_income = as.numeric(med_income)) |>
  select(GEOID, med_income)|>
  filter(!is.na(med_income))

merged <- merged |>
  left_join(INCOME, by = "GEOID") 

# ---- HEALTH ----
library(httr)
library(rvest)
library(jsonlite)

places_cache_file <- file.path("data", "final", "places_data_cache.csv")

if (file.exists(places_cache_file)) {

  places_data <- read_csv(places_cache_file)

} else {
  cat("No cache found. Downloading PLACES data from CDC...\n")

  download_places_data <- function(measure, limit = 50000, offset = 0) {
    url <- "https://data.cdc.gov/resource/cwsq-ngmh.json"
    
    query_url <- paste0(
      url,
      "?$limit=", limit,
      "&$offset=", offset,
      "&measureid=", measure
    )
    
    response <- GET(query_url)
    
    if (status_code(response) == 200) {
      data <- fromJSON(content(response, "text", encoding = "UTF-8"))
      return(data)
    } else {
      warning(paste("Error downloading data for", measure, ":", status_code(response)))
      return(NULL)
    }
  }

  cat("Downloading diabetes data...\n")
  diabetes_data <- list()
  offset <- 0
  chunk_size <- 50000
  
  repeat {
    cat(paste("  Downloading records starting at", offset, "...\n"))
    chunk <- download_places_data("DIABETES", limit = chunk_size, offset = offset)
    
    if (is.null(chunk)) {
      cat("  No more data or error occurred\n")
      break
    }
    if (nrow(chunk) == 0) break
    
    diabetes_data[[length(diabetes_data) + 1]] <- chunk
    
    if (nrow(chunk) < chunk_size) break
    
    offset <- offset + chunk_size
  }
  
  cat("Downloading obesity data...\n")
  obesity_data <- list()
  offset <- 0
  
  repeat {
    cat(paste("  Downloading records starting at", offset, "...\n"))
    chunk <- download_places_data("OBESITY", limit = chunk_size, offset = offset)
    
    if (is.null(chunk)) {
      cat("  No more data or error occurred\n")
      break
    }
    if (nrow(chunk) == 0) break
    
    obesity_data[[length(obesity_data) + 1]] <- chunk
    
    if (nrow(chunk) < chunk_size) break
    
    offset <- offset + chunk_size
  }

  cat("Downloading food insecurity data...\n")
  food_insecurity_data <- list()
  offset <- 0
  
  repeat {
    cat(paste("  Downloading records starting at", offset, "...\n"))
    chunk <- download_places_data("FOODINSECURITY", limit = chunk_size, offset = offset)
    
    if (is.null(chunk) || length(chunk) == 0) {
      cat("  Food insecurity data not available or no more data\n")
      break
    }
    if (!is.data.frame(chunk) || nrow(chunk) == 0) break
    
    food_insecurity_data[[length(food_insecurity_data) + 1]] <- chunk
    
    if (nrow(chunk) < chunk_size) break
    
    offset <- offset + chunk_size
  }
  
  cat("Downloading SNAP/food stamps data...\n")
  snap_data <- list()
  offset <- 0
  
  repeat {
    cat(paste("  Downloading records starting at", offset, "...\n"))
    chunk <- download_places_data("FOODSTAMPS", limit = chunk_size, offset = offset)
    
    if (is.null(chunk) || length(chunk) == 0) {
      cat("  SNAP data not available or no more data\n")
      break
    }
    if (!is.data.frame(chunk) || nrow(chunk) == 0) break
    
    snap_data[[length(snap_data) + 1]] <- chunk
    
    if (nrow(chunk) < chunk_size) break
    
    offset <- offset + chunk_size
  }

  cat("Downloading high blood pressure data...\n")
  bphigh_data <- list()
  offset <- 0
  
  repeat {
    cat(paste("  Downloading records starting at", offset, "...\n"))
    chunk <- download_places_data("BPHIGH", limit = chunk_size, offset = offset)
    
    if (is.null(chunk) || length(chunk) == 0) {
      cat("  High blood pressure data not available or no more data\n")
      break
    }
    if (!is.data.frame(chunk) || nrow(chunk) == 0) break
    
    bphigh_data[[length(bphigh_data) + 1]] <- chunk
    
    if (nrow(chunk) < chunk_size) break
    
    offset <- offset + chunk_size
  }
  
  cat("Downloading high cholesterol data...\n")
  cholesterol_data <- list()
  offset <- 0
  
  repeat {
    cat(paste("  Downloading records starting at", offset, "...\n"))
    chunk <- download_places_data("HIGHCHOL", limit = chunk_size, offset = offset)
    
    if (is.null(chunk) || length(chunk) == 0) {
      cat("  High cholesterol data not available or no more data\n")
      break
    }
    if (!is.data.frame(chunk) || nrow(chunk) == 0) break
    
    cholesterol_data[[length(cholesterol_data) + 1]] <- chunk
    
    if (nrow(chunk) < chunk_size) break
    
    offset <- offset + chunk_size
  }
  
  places_data <- bind_rows(c(diabetes_data, obesity_data, food_insecurity_data, 
                             snap_data, bphigh_data, cholesterol_data))
  
  places_data <- places_data %>%
    select(where(~ !is.list(.)))
  
  cat(paste("Downloaded", nrow(places_data), "records\n"))
  
  write_csv(places_data, places_cache_file)
  
}

if ("tractfips" %in% names(places_data)) {
  tract_col <- "tractfips"
} else if ("locationid" %in% names(places_data)) {
  tract_col <- "locationid"
} else if ("tract_fips" %in% names(places_data)) {
  tract_col <- "tract_fips"
} else {
  stop("Cannot find tract identifier column. Available columns: ", paste(names(places_data), collapse=", "))
}

health_data <- places_data %>%
  filter(measureid %in% c("DIABETES", "OBESITY", "FOODINSECURITY", "FOODSTAMPS", 
                          "BPHIGH", "HIGHCHOL")) %>%
  select(any_of(c("locationname", "stateabbr", "statedesc", tract_col, "measureid", 
                  "data_value", "totalpopulation", "total_population"))) %>%
  rename(tractfips = all_of(tract_col))

if ("totalpopulation" %in% names(health_data)) {
  health_data <- health_data %>%
    mutate(
      data_value = as.numeric(data_value),
      totalpopulation = as.numeric(totalpopulation)
    )
} else if ("total_population" %in% names(health_data)) {
  health_data <- health_data %>%
    mutate(
      data_value = as.numeric(data_value),
      totalpopulation = as.numeric(total_population)
    )
} else {
  health_data <- health_data %>%
    mutate(
      data_value = as.numeric(data_value),
      totalpopulation = NA_real_
    )
}

health_wide <- health_data %>%
  select(tractfips, stateabbr, measureid, data_value, totalpopulation, any_of("locationname")) %>%
  distinct() %>%
  pivot_wider(
    names_from = measureid,
    values_from = data_value,
    id_cols = c(tractfips, stateabbr, totalpopulation, any_of("locationname"))
  ) %>%
  rename_with(tolower)

if ("locationname" %in% names(health_wide)) {
  health_wide <- health_wide %>%
    mutate(location = locationname)
} else {
  health_wide <- health_wide %>%
    mutate(location = NA_character_)
}

write_csv(health_wide, file.path("data", "final", "health_data_processed.csv"))

#cat("Success! Processed data saved to locally")
```

### Linear Regression: Diabetes
```{r}
#| message: false
#| warning: false
DIABETES <- health_wide |>
  rename('GEOID' = 'tractfips') |>
  select(GEOID, diabetes)

MERGED_DIABETES <- merged |>
  left_join(DIABETES, by = "GEOID") 

# ---- Linear Regression for Diabetes ---- 
library(ggplot2)
library(ggpmisc)

model_data_dia <- MERGED_DIABETES |>
  select(diabetes, med_income, den_fastfood, LA1and10) |>
  na.omit()

model_dia <- lm(diabetes ~ med_income + den_fastfood + LA1and10, data = model_data_dia)

model_data_dia$predicted <- predict(model_dia)

ggplot(model_data_dia, aes(x = diabetes, y = predicted)) +
  geom_point(alpha = 0.2, color = "#9badff") +
  geom_abline(slope = 1, intercept = 0, color = "#0432ff", linetype = "dashed", size = 1) +
  stat_poly_eq() + 
  labs(title = "Actual vs Predicted Diabetes Prevalence",
       subtitle = "Linear regression using income, fast food density,\nand food desert variables",
       x = "Actual Diabetes Rate(%)",
       y = "Predicted Diabetes Rate(%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 9))
```

```{r}
# Tidy the model coefficients
tidy_model <- tidy(model_dia)

# Format the numbers for presentation
presentation_data <- tidy_model |>
  mutate(
    # Rename 'term' column for presentation
    Predictor = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "med_income" ~ "Median Income",
      term == "den_fastfood" ~ "Fast Food Density",
      term == "LA1and10" ~ "Low Access (1 & 10 mi)",
      TRUE ~ term
    ),
    # Round or format columns
    Estimate = round(estimate, 5),
    `p-value` = format.pval(p.value, digits = 3, eps = 0.001) # Better p-value formatting
  ) |>
  # Select and reorder columns
  select(Predictor, Estimate, `p-value`)

datatable(
  presentation_data,
  caption = "Linear Regression Model Results for Diabetes Prevalence",
  # Customize the display options
  options = list(
    dom = 't', # Only show the table (no search, pagination, etc.)
    pageLength = nrow(presentation_data), # Show all rows
    autoWidth = TRUE
  ),
  rownames = FALSE # Remove the R row numbers
)
```

### Linear Regression: Obesity
```{r}
OBESITY <- health_wide |>
  rename('GEOID' = 'tractfips') |>
  select(GEOID, obesity)

MERGED_OBESITY <- merged |>
  left_join(OBESITY, by = "GEOID")  

# ---- Linear Regression for Obesity ---- 

model_data_obesity <- MERGED_OBESITY |>
  select(obesity, med_income, den_fastfood, LA1and10) |>
  na.omit()

model_obesity <- lm(obesity ~ med_income + den_fastfood + LA1and10, data = model_data_obesity)

model_data_obesity$predicted <- predict(model_obesity)

ggplot(model_data_obesity, aes(x = obesity, y = predicted)) +
  geom_point(alpha = 0.2, color = "#9badff") +
  geom_abline(slope = 1, intercept = 0, color = "#0432ff", linetype = "dashed", size = 1) +
  stat_poly_eq() + 
  labs(title = "Actual vs Predicted Obesity Prevalence",
       subtitle = "Linear regression using income, fast food density,\nand food desert variables",
       x = "Actual Obesity Rate(%)",
       y = "Predicted Obesity Rate(%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 9))
```

```{r}
tidy_model_o <- tidy(model_obesity)

# 2. Format the numbers for presentation
presentation_data_o <- tidy_model_o |>
  mutate(
    # Rename 'term' column for presentation
    Predictor = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "med_income" ~ "Median Income",
      term == "den_fastfood" ~ "Fast Food Density",
      term == "LA1and10" ~ "Low Access (1 & 10 mi)",
      TRUE ~ term
    ),
    # Round or format columns
    Estimate = round(estimate, 5),
    `p-value` = format.pval(p.value, digits = 3, eps = 0.001) # Better p-value formatting
  ) |>
  # Select and reorder columns
  select(Predictor, Estimate, `p-value`)

# Print the interactive table
datatable(
  presentation_data_o,
  caption = "Linear Regression Model Results for Obesity Prevalence",
  # Customize the display options
  options = list(
    dom = 't', # Only show the table (no search, pagination, etc.)
    pageLength = nrow(presentation_data), # Show all rows
    autoWidth = TRUE
  ),
  rownames = FALSE # Remove the R row numbers
)

```

### Linear Regression: Obesity (No Low Access)
```{r}
model_data_obesity_v2 <- MERGED_OBESITY |>
  select(obesity, med_income, den_fastfood) |>
  na.omit()

# Create the Linear Regression Model
model_obesity_v2 <- lm(obesity ~ med_income + den_fastfood, data = model_data_obesity_v2)

model_data_obesity_v2$predicted <- predict(model_obesity_v2)

ggplot(model_data_obesity_v2, aes(x = obesity, y = predicted)) +
  geom_point(alpha = 0.2, color = "#D86ECC") +
  geom_abline(slope = 1, intercept = 0, color = "#0432ff", linetype = "dashed", size = 1) +
  stat_poly_eq() + 
  labs(title = "Actual vs Predicted Obesity Prevalence",
       subtitle = "Linear regression only using income and fast food density",
       x = "Actual Obesity Rate(%)",
       y = "Predicted Obesity Rate(%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 9))
```

```{r}
tidy_model_o2 <- tidy(model_obesity_v2)

presentation_data_o2 <- tidy_model_o2 |>
  mutate(
    Predictor = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "med_income" ~ "Median Income",
      term == "den_fastfood" ~ "Fast Food Density",
      term == "LA1and10" ~ "Low Access (1 & 10 mi)",
      TRUE ~ term
    ),
    Estimate = round(estimate, 5),
    `p-value` = format.pval(p.value, digits = 3, eps = 0.001) # Better p-value formatting
  ) |>
  select(Predictor, Estimate, `p-value`)

datatable(
  presentation_data_o2,
  caption = "Linear Regression Model Results for Obesity Prevalence",
  # Customize the display options
  options = list(
    dom = 't', # Only show the table (no search, pagination, etc.)
    pageLength = nrow(presentation_data), # Show all rows
    autoWidth = TRUE
  ),
  rownames = FALSE # Remove the R row numbers
)

```




