---
title: "NYC Trees"
format: 
  html:
    theme: morph
    css: custom.css
    toc: true
    toc-location: right
    toc-depth: 3
    code-fold: true
    list:
      indent: false
---

# Executive Summary 
[insert text]

# Data Acquisition

### NYC City Council District Boundaries
```{r}
#| message: false
#| warning: false

#library function to install package if needed
library <- function(pkg){
  pkg <- as.character(substitute(pkg))
  options(repos = c(CRAN = "https://cloud.r-project.org"))
  if(!require(pkg, character.only=TRUE, quietly=TRUE)) install.packages(pkg)
  stopifnot(require(pkg, character.only=TRUE, quietly=TRUE))
}

library(httr2)
library(dplyr)
library(readr)
library(ggplot2)
library(DT)
library(stringr)
library(sf)
library(jsonlite)

if(!dir.exists(file.path("data", "mp03"))){
  dir.create(file.path("data", "mp03"), showWarnings=FALSE, recursive=TRUE)
}

NYC_DISTRICT_BOUNDARIES_FILENAME <- file.path("data", "mp03", "nycc_25c.zip")

if(!file.exists(NYC_DISTRICT_BOUNDARIES_FILENAME)){
  download.file("https://s-media.nyc.gov/agencies/dcp/assets/files/zip/data-tools/bytes/city-council/nycc_25c.zip", 
                destfile=NYC_DISTRICT_BOUNDARIES_FILENAME)
}

unzip(NYC_DISTRICT_BOUNDARIES_FILENAME, exdir = file.path("data", "mp03"))

#calling for unzipped folder
NYC_DISTRICT_BOUNDARIES_UNZ_FILENAME <- file.path("data", "mp03", "nycc_25c")


#step 4: reading file 
nyc_boundaries <- st_read(file.path(NYC_DISTRICT_BOUNDARIES_UNZ_FILENAME, 
                                   "nycc.shp"), quiet = TRUE)

#crs= coordinate reference system
nyc_boundaries <- st_transform(nyc_boundaries, crs="WGS84")
```

### NYC Tree Points

```{r}
#| message: false
#| warning: false

ENDPOINT <- "https://data.cityofnewyork.us/resource/hn5i-inap.json"

BATCH_SIZE <- 50000
OFFSET     <- 0
END_OF_EXPORT <- FALSE
ALL_DATA <- list()

while(!END_OF_EXPORT){
  cat("Requesting items", OFFSET, "to", BATCH_SIZE + OFFSET, "\n")
  
  req <- request(ENDPOINT) |>
    req_url_query(`$limit`  = BATCH_SIZE, 
                  `$offset` = OFFSET)
  
  resp <- req_perform(req)
  
  batch_data <- fromJSON(resp_body_string(resp))
  
  ALL_DATA <- c(ALL_DATA, list(batch_data))
  
  if(NROW(batch_data) != BATCH_SIZE){
    END_OF_EXPORT <- TRUE
    
    cat("End of Data Export Reached\n")
  } else {
    OFFSET <- OFFSET + BATCH_SIZE
  }
}

ALL_DATA <- bind_rows(ALL_DATA)
cat("Data export complete:", NROW(ALL_DATA), "rows and", NCOL(ALL_DATA), "columns.")

#saving as a rds file
saveRDS(ALL_DATA, file = "data/mp03/nyc_trees_export.rds")
cat("Data successfully saved to data/mp03/nyc_trees_export.rds\n")

#reading nyc_trees_export.rds
nyc_trees <- read_rds(file = "data/mp03/nyc_trees_export.rds")
```


# Data Integration and Initial Exploration

### Plot All Tree Points
```{r}
#| message: false
#| warning: false

#changing the geometry column into spatial data 
nyc_trees <- st_as_sf(nyc_trees, wkt = "geometry", crs = "WGS84")

#plotting all the trees in nyc
ggplot() +
  geom_sf(data = nyc_boundaries, fill = "white", color = "black", linewidth = 0.5) +
  geom_sf(data = nyc_trees, color = "forestgreen", alpha = 0.07, size = 0.3) +
  theme_bw() +
  labs(title = "NYC Trees within Council Districts",
       subtitle = "NYC District Shoreline Boundaries ",
       caption = "Source: NYC Open Data")
```

### District-Level Analysis of Tree Coverage

Joining the Trees and NYC City Council District datasets

```{r}
tree_districts <- st_join(
  nyc_trees, nyc_boundaries, join = st_intersects)
```

1. Which council district has the most trees?

```{r}
top_tree_count <- tree_districts |>
  group_by(CounDist)|>
  summarise(total_trees_per_area = n())|>
  slice_max(total_trees_per_area, n = 1)|>
  pull(CounDist)
```

**`r top_tree_count`** District has the most trees. 

2. Which council district has the highest density of trees? 

```{r}
highest_density <- tree_districts |>
  group_by(CounDist) |>
  summarise(
    total_trees_per_area = n(),
    tree_density = total_trees_per_area / first(Shape_Area)) |>
  slice_max(tree_density, n = 1) |>
  pull(CounDist)
```

**`r highest_density`** District has the highest density of trees. 

3. Which district has highest fraction of dead trees out of all trees?

```{r}
most_dead_ratio <- tree_districts |>
  group_by(CounDist) |>
  summarise(all = n(),
            dead = sum(tpcondition == "Dead", na.rm = TRUE)) |>
  mutate(fraction = dead/all) |>
  arrange(desc(fraction)) |>
  slice_head(n=1)|>
  pull(CounDist)
```
**`r most_dead_ratio`** District had the highest fraction of dead trees out of all trees. 

4. What is the most common tree species in Manhattan?

```{r}
top_tree_manhattan <- tree_districts |>
  mutate( 
    borough = case_when(
      between(CounDist, 1,10) ~ 'Manhattan',
      between(CounDist, 11,18) ~ 'Bronx',
      between(CounDist, 19,32) ~ 'Queens',
      between(CounDist, 33,48) ~ 'Brooklyn',
      between(CounDist, 49,51) ~ 'Staten Island'
    ))|>
  filter(borough == 'Manhattan')|>
  group_by(genusspecies)|>
  summarise(species_count = n())|>
  slice_max(species_count)|>
  pull(genusspecies)
```

The most common tree in Manhattan is **`r top_tree_manhattan`**.

5. What is the species of the tree closest to Baruch’s campus?

```{r}
#function to create a location point
new_st_point <- function(lat, lon, ...){
  st_sfc(point = st_point(c(lon, lat))) |>
    st_set_crs("WGS84")
}

baruch_point <- new_st_point(40.7403, -73.9833)

closest_tree <- tree_districts |>
  mutate(distance = st_distance(geometry, baruch_point))|>
  select(CounDist,distance, genusspecies)|>
  slice_min(distance, n=1) |>
  pull(genusspecies)
```
The closest tree to the Baruch Campus is a **`r closest_tree`**.

# NYC Tree Project Proposal




